---
title: "Day3: nonlinear models"
description: |
  Nature has no straight lines
author:
  - name: Andrew MacDonald
    url: {}
date: 2022-03-29
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Fitting nonlinear models to ecological data is interesting and powerful. 
This is possible in base R using the function `nls()`. 
In a Bayesian approach we can do the same thing, but we don't need to learn any new tools.


## Hemlock growth

we will work with a dataset 

```{r}
library(tidyverse)
library(tidybayes)
library(brms)

hemlock <- readr::read_delim(
  "https://raw.githubusercontent.com/bios2/biodiversity_modelling_2021/master/data/hemlock.txt", 
  delim = " ",
  col_names = c("x","light", "growth"), skip = 1)

knitr::kable(head(hemlock, n = 3))



ggplot(hemlock, aes(x = light, y = growth)) + 
  geom_point()

```


## define a model

We need a function for the mean growth rate per species. 
A very popular choice in ecology is the famous **Type 2 functional response**:

$$
y = \frac{a x}{b + x}
$$
* $a$ is the asymptote -- the max value of $y$ when $x$ is large
* $b$ is the value of $x$ where $y = a/2$

We experiment using `curve` to understand how this works

```{r}
a <- 195
b <- 30
curve(a * x / (b + x), xlim = c(0, 100))
```


## define a distribution for observations around this average


We can use the gamma distribution. The gamma distribution looks like this:

```{r}
curve(dgamma(x, 3,5), xlim = c(0, 3))
```

The gamma distribution has two parameters:

$$
\text{Gamma}(a, b)
$$

The mean and variance are both functions of both of these parameters:

$$
\mu = \frac{a}{b}
$$
$$
\sigma^2 = \frac{a}{b^2}
$$

we can demonstrate this easily in R:

```{r}
xx <- rgamma(5000, 3, 5)
mean(xx) #about 3/5  = 0.6
var(xx) #about 3/(5^2) = 0.12
```

We can reverse this: write the parameters $a$ and $b$ in terms of the desired mean and standard deviation:

$$
\begin{align}
a &= \frac{\mu^2}{\sigma^2} \\
b &= \frac{\mu}{\sigma^2} \\
\end{align}
$$

_optional_ prove that to yourself with algebra!

*exercise* simulate 3000 from a Gamma distribution with a mean of 42 and a standard deviation of 10

### simulating observations:

we exploit this technique to make up some fake data:

```{r}
a <- 195
b <- 20
x_values <- runif(n = 70, min = 0, max = 100)
average_response <- a * x_values / (b + x_values)
plot(x_values, average_response)

```

```{r}
sigma <- 31

observed_response <- rgamma(n = 70, shape = average_response^2/sigma^2, rate = average_response/ sigma^2)

plot(x_values, observed_response)
```

## Defining a bayesian model

To fully build our Bayesian model we put all the above together:

* our function that describes how light causes the _average_ growth
* a distribution for observations around that average
* priors for the three unobserved quantities: $a$, $b$ and $\sigma$

$$
\begin{align}
\text{growth} &\sim \text{Gamma}(\mu^2/\sigma^2, \mu/\sigma^2) \\
\mu &= \frac{aL}{b + L} \\
\sigma & \sim \text{Exponential}(4)\\
a               & \sim \text{Normal}(200, 15)\\
b               & \sim \text{Normal}(25, 5)\\
\end{align}
$$
## Prior predictive simulations

```{r}
hemlock$x <- NULL

light_curve_bf <- bf(growth ~ exp(loga) * light / (exp(logb) + light),
                        family = Gamma(link = "identity"),
                        loga ~ 1,
                        logb ~ 1, 
                        nl = TRUE)

get_prior(light_curve_bf, data = hemlock)

light_curve_prior <- c(
  # prior(exponential(.1), class = "shape"),
  prior(gamma(6.25, .25), class = "shape"),
  prior(gamma(250^2/70^2, 250/70^2), class = "b", nlpar = "a", lb = 0),
  prior(normal(30, 20), class = "b", nlpar = "b")
)

light_curve_model_prior <- brm(light_curve_bf,
                               prior = light_curve_prior,
                               data = hemlock,
                               refresh = FALSE,
                               sample_prior = "only", 
                               file = here::here("_posts", "2022-03-29-day3-nonlinear-models", "light_curve_prior"),
                               file_refit = "on_change")
```


```{r}

hemlock |> 
  add_predicted_draws(light_curve_model_prior, ndraws = 6) |> 
  ggplot(aes(x = light, y = .prediction)) + 
  geom_point() + 
  facet_wrap(~.draw) + 
  coord_cartesian(ylim = c(0, 300))
```

fit to real data:

```{r}
light_curve_model_posterior <- brm(light_curve_bf,
                               prior = light_curve_prior,
                               data = hemlock,
                               sample_prior = "yes", 
                               file = here::here("_posts", "2022-03-29-day3-nonlinear-models", "light_curve_posterior"),
                               file_refit = "on_change")
```

Predictions with the original data:

```{r}

hemlock_post <- hemlock |> 
  add_predicted_draws(light_curve_model_posterior)


hemlock_post |> 
  ggplot(aes(x = light, y = .prediction)) + 
  stat_lineribbon()+ 
  coord_cartesian(ylim = c(0, 300)) + 
  scale_fill_brewer(palette = "Oranges") + 
  geom_point(aes(x = light, y = growth), size = 3, pch = 21,
             fill = "lightblue", data = hemlock)
```

### multilevel nonlinear model


```{r}




light_curve_bf <- bf(growth ~ exp(loga) * light / (exp(logb) + light),
                        family = Gamma(link = "identity"),
                        loga ~ 1 + (1|ben|spp),
                        logb ~ 1 + (1|ben|spp), 
                        nl = TRUE)

get_prior(light_curve_bf, data = hemlock |> mutate(spp = "a"))

light_curve_prior <- c(
  # prior(exponential(.1), class = "shape"),
  prior(gamma(6.25, .25), class = "shape"),
  prior(normal(5.5, 3), class = "b", nlpar = "loga"),
  prior(normal(3.4, 2), class = "b", nlpar = "logb"),
  prior(exponential(8), class = "sd", group = "spp", nlpar = "loga"),
  prior(exponential(8), class = "sd", group = "spp", nlpar = "logb"),
  prior(lkj(2), class = "cor")
)

fake_hemlock <- expand.grid(spp = letters[1:3],
                            light = runif(120, min = 0, max = 100), 
                            growth = 1)

light_curve_model_prior <- brm(light_curve_bf,
                               prior = light_curve_prior,
                               data = fake_hemlock,
                               refresh = FALSE,
                               sample_prior = "only", 
                               file = here::here("_posts", "2022-03-29-day3-nonlinear-models", "light_curve_prior"),
                               file_refit = "on_change")
```

```{r}
fake_spp_prior <- fake_hemlock |> 
  add_predicted_draws(light_curve_model_prior, ndraws = "4")


fake_spp_prior |> 
  ggplot(aes(x = light , y = .prediction)) + geom_point() + 
  facet_grid(spp~.draw)
```

