---
title: "day3-offsets"
description: |
  A short description of the post.
author:
  - name: Andrew MacDonald
    url: {}
date: 2022-03-29
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

let's imagine that we have counted dandelions. 

Dandelions occur on average 6 per square meter

However we have five kinds of quadrat: 1, 4, 9 and 25 square meters

```{r}
library(tidyverse)

imaginary_dandelions <- tibble(quadrat_size = rep(c(1,4, 9, 25), each = 15),
       n_per_m2 = purrr::map(quadrat_size, rpois, lambda = 6),
       obs_dandelions = map_dbl(n_per_m2, sum))

ggplot(imaginary_dandelions, aes(x = obs_dandelions)) + geom_histogram() + 
  facet_wrap(~quadrat_size)

```

How can we get the correct number of dandelions? 


## Poisson count model

$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) &= \beta
\end{align}
$$
$\lambda$ is the average response. If we want to measure the average _per unit effort_, we can do that too:

$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda/Q) &= \beta
\end{align}
$$




$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) - \text{log}(Q) &= \beta
\end{align}
$$


$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) &= \beta + \text{log}(Q)
\end{align}
$$

In other words, we need a way to add a log coefficient to a model and give it a slope of _exactly one_. Fortunately the function `offset()` is here to do exactly this:

```{r}
dandelion_model <- glm(obs_dandelions ~ 1, family = poisson(link = "log"), data = imaginary_dandelions)
summary(dandelion_model) 

```

This gives the wrong answer! 
 
 
```{r}
dandelion_model <- glm(obs_dandelions ~ 1 + offset(log(quadrat_size)),
                       family = poisson(link = "log"),
                       data = imaginary_dandelions)
summary(dandelion_model) 
```


the coefficient should be close to 6, after we reverse the link function:

```{r}
exp(coef(dandelion_model)[[1]])
```
 
