---
title: "Day3: offsets"
description: |
  How to do a count analysis for different efforts, or exposure.
author:
  - name: Andrew MacDonald
    url: {}
date: 2022-03-29
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

let's imagine that we have counted dandelions. 

Dandelions occur on average 6 per square meter

However we have five kinds of quadrat: 1, 4, 9 and 25 square meters

```{r}
library(tidyverse)

imaginary_dandelions <- tibble(quadrat_size = rep(c(1,4, 9, 25), each = 15),
       n_per_m2 = purrr::map(quadrat_size, rpois, lambda = 6),
       obs_dandelions = map_dbl(n_per_m2, sum))

ggplot(imaginary_dandelions, aes(x = obs_dandelions)) + geom_histogram() + 
  facet_wrap(~quadrat_size)

```

How can we get the correct number of dandelions? 


## Poisson count model

$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) &= \beta
\end{align}
$$
$\lambda$ is the average response. If we want to measure the average _per unit effort_, we can do that too:

$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda/Q) &= \beta
\end{align}
$$




$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) - \text{log}(Q) &= \beta
\end{align}
$$


$$
\begin{align}
y &\sim \text{Poisson}(\lambda) \\
\text{log}(\lambda) &= \beta + \text{log}(Q)
\end{align}
$$

In other words, we need a way to add a log coefficient to a model and give it a slope of _exactly one_. Fortunately the function `offset()` is here to do exactly this:

```{r}
dandelion_model <- glm(obs_dandelions ~ 1, family = poisson(link = "log"), data = imaginary_dandelions)
summary(dandelion_model) 

```

This gives the wrong answer! 
 
 
```{r}
dandelion_model <- glm(obs_dandelions ~ 1 + offset(log(quadrat_size)),
                       family = poisson(link = "log"),
                       data = imaginary_dandelions)
summary(dandelion_model) 
```


the coefficient should be close to 6, after we reverse the link function:

```{r}
exp(coef(dandelion_model)[[1]])
```


## Do it the Bayes Way

```{r}
library(brms)

dandelion_bf <- bf(obs_dandelions ~ 1 + offset(log(quadrat_size)), 
                   family = poisson(link = "log"))

get_prior(dandelion_bf, data = imaginary_dandelions)

dandelion_prior <- prior(normal(2, 1), class = "Intercept")

dandelion_model <- brm(formula = dandelion_bf,
                       data = imaginary_dandelions, 
                       prior = dandelion_prior)

```

look at stan code

```{r}
stancode(dandelion_model)
```

look at posterior distribution of parameter

```{r}
# as.matrix(dandelion_model) |> head()

library(tidybayes)

tidy_draws(dandelion_model) |> 
  ggplot(aes(x = exp(b_Intercept))) + 
  geom_histogram() + 
  geom_vline(xintercept = 6, col = "red", lwd = 3)
```

